What we {\sf eval} in the shadows

### 2021-04-01 paper/main.tex
5.1

\begin{itemize}
\item \underline{$min(e)=\c{V}$}: Expression that represent values are
  frequently passed to \eval. The majority of those, \packageValOneNodePercent,
  are inline constants (integers or doubles). The rest trivially evaluate to a
  value.\footnote{True as long as base functions such as \c{+} are not
  redefined. They typically aren't, but this is a limitation of this
  categorization.} In our corpus, \packageNbCallSitesUniqueActualValue call
  sites only ever see a simple values; the median number of times they are
  exercised is \packageMedianRunSitesUniqueActualValue, suggesting a possible
  lack of coverage. However, manual inspection reveals cases such as:
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
  This concatenates to a string and \eval simply returns it as is.
  From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
  them needed to use \eval.
%% The percentage of non default environment passed to \eval is
%% \packageMinimizedpercentenvirc when the minimized expression is a value. Some
%% sites which are passed a value, are also passed different minimized expressions.
%% For those ones, \packageNonDefaultEnvirValuePercent use a non-default
%% environment, and among the one that use a non-default environment,
%% \packageNonDefaultEnvirWithVarPercent have at least one minimized expression
%% with a variable inside.

\item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
  subsume values, thus \c{X} includes arithmetic expressions in which a variable
  occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
  just an \c{x}. As a variable may refer to a promise, it may trigger
  computation.
  %% The
  %% \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of
  %% the call sites. The call sites without an explicit \c{envir} argument all use
  %% a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
  %% both explicit and different from \c{parent.frame()} in
  %% \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that
  %% when given a variable, \eval is often used to get value in a non-local,
  %% non-parent environment.
\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, can be used to
  access variable in an environment or a column in a data frame, this also
  include vector   indexing operators, \c{[]} and \c{[[]]}. A name lookup
  subsumes variables.
%% It amounts for \packageNbSlotAccess sites, \ie
%% \packageSlotAccessPercent of the sites. The non-default, non-parent environment
%% percentage is here \packageMinimizedpercentenvirnonparentg.

\item \underline{$min(e)=$~\c{<-}}: Assignments account for
  \packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
  \c{assign}, and they subsume named lookups.
%% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
%% most of those assignments occur in the default environment, \ie in the caller of
%% \eval. This hints that side effects are not widespread in \eval.\footnote{Side
%% effects can also happen in a function call in the \eval expression, and that
%% function call will hide assignment operators from us. We more precisely
%% instrument side-effects to environment in Section~REF.}

\item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
  values as argument but does not include variables. It should be noted that
  function name require variable lookups -- we distinguish those as they usually
  do not trigger computation.

\item \underline{$min(e)=\c{F(X)}$}: Function calls whose arguments may include
  variable references.

@@

\item \underline{$min(e)=\c{FUN}$}: Function definition accounts for
  \packageFunctionDefinitionSitesPercent of \eval sites. In addition the
  functions inside other expressions account for
  \packageGeneralizedFunctionDefinitionSitesPercent of sites, and give an
  estimate of the use of higher-order functions.

\item \underline{$min(e)=\c{\{\,\}}$}: Arbitrary-length statement blocks. These
  are larger expressions with, on average, \packageMinimizedoperationsjRnd
  instructions.

\item \underline{$min(e)=\c{model.frame}$}: Call to the \c{model.frame} function
  used to create statistical models.

@@

Similarly, when a function is evaluated on some variable (\c{F(X)} minimized
expression), the percentage of non default, non parent environments is rather
high, \packageMinimizedpercentenvirnonparentd\%.

As a whole, variable access, either as the top call, or inside a function call,
possibly with accessing a slot, represents
\packageGeneralizedVarAccessSitePercent of the sites.

The minimized expression classification suggests that the presence of variables
in the expression is highly correlated to the specification of a specific
environment, and that \eval is used here for its power in precising the
evaluation environment.

Table~\ref{tab:minimizedexpressions} gives eleven most frequent in number of
sites. \#sites is the number of sites; \%sites the percentage of sites across
the total number of sites (\packageNbCallSites sites) with at least this
minimized expression; \#packages is the number of packages with at least one of
the minimized expression and measures how widespread a minimized expression is.
\#operations is the average number of steps performed by the R interpreter when
given that minimized expression, and \%envir is the percentage of call sites for
which the \c{envir} argument is explicitly passed, and is not
\c{parent.frame()}, which is the most common explicit environment passed to
\eval.

@@

Note that a site can exhibit several minimized expressions (see
Figure~\ref{fig:frequentcombinationminimized}). However,
\packageNbOneMinimizedPercent of the sites have only one minimized expression.
Note also that if a function is called in that expression, we see only the
function call but not the body of the function. We will trace what happens
during the evaluation of the expression by \eval in
Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.

\NOTE{Maybe we should ditch the table of the minimized expressions and rather
  use the ones of combinations. It has the advantage of showing what is the most
  frequent, per call site. But it's a long tail, and sites that have several
  minimized expressions are not that numerous, only 13\%. But if include
  combinations, percentages change. For instance, 3314 call sites have only X,
  i.e. 22.51, instead of 4353 and 29.6\%. }

\paragraph{Base}
The ranking of minimized expressions in \emph{base} is similar to the one in
CRAN but  the share of \c{V} is much larger, as it is present in
\baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in
\baseMinimizedpropsitesb\% of the sites.

\paragraph{Kaggle}
Kaggle is dominated by sites with minimized expression \c{V}, with
\kaggleMinimizedcallsitesa sites, and \kaggleMinimizedpropsitesa\% of the
\kaggleNbCallSites sites in total. All of those values are an AST size of one
node, suggesting they do not come. \NOTE{TODO}
%TODO: all the src ref here seem to be incorrect and not from kaggle.

\subsubsection{The environment passed to \eval.}

The \c{envir} argument of \eval can be an actual environment, but also \c{Null}, a list, a dataframe, a pairlist or even an integer. % Sometimes also a double, but it is cast to integer anyway.

>>>>

\begin{itemize}
\item \underline{$min(e)=\c{V}$}: Expression that represent values are
  frequently passed to \eval. The majority of those, \packageValOneNodePercent,
  are inline constants (integers or doubles). The rest trivially evaluate to a
  value.\footnote{True as long as base functions such as \c{+} are not
  redefined. They typically aren't, but this is a limitation of this
  categorization.} In our corpus, \packageNbCallSitesUniqueActualValue call
  sites only ever see a simple values; the median number of times they are
  exercised is \packageMedianRunSitesUniqueActualValue, suggesting a possible
  lack of coverage. However, manual inspection reveals cases such as:
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
  This concatenates to a string and \eval simply returns it as is.
  From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
  them needed to use \eval.
%% The percentage of non default environment passed to \eval is
%% \packageMinimizedpercentenvirc when the minimized expression is a value. Some
%% sites which are passed a value, are also passed different minimized expressions.
%% For those ones, \packageNonDefaultEnvirValuePercent use a non-default
%% environment, and among the one that use a non-default environment,
%% \packageNonDefaultEnvirWithVarPercent have at least one minimized expression
%% with a variable inside.
\item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
  subsume values, thus \c{X} includes arithmetic expressions in which a variable
  occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
  just an \c{x}. As a variable may refer to a promise, it may trigger
  computation.
  %% The
  %% \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of
  %% the call sites. The call sites without an explicit \c{envir} argument all use
  %% a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
  %% both explicit and different from \c{parent.frame()} in
  %% \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that
  %% when given a variable, \eval is often used to get value in a non-local,
  %% non-parent environment.
\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, can be used to
  access variable in an environment or a column in a data frame, this also
  include vector   indexing operators, \c{[]} and \c{[[]]}. A name lookup
  subsumes variables.
%% It amounts for \packageNbSlotAccess sites, \ie
%% \packageSlotAccessPercent of the sites. The non-default, non-parent environment
%% percentage is here \packageMinimizedpercentenvirnonparentg.
\item \underline{$min(e)=$~\c{<-}}: Assignments account for
  \packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
  \c{assign}, and they subsume named lookups.
%% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
%% most of those assignments occur in the default environment, \ie in the caller of
%% \eval. This hints that side effects are not widespread in \eval.\footnote{Side
%% effects can also happen in a function call in the \eval expression, and that
%% function call will hide assignment operators from us. We more precisely
%% instrument side-effects to environment in Section~REF.}
\item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
  values as argument but does not include variables. It should be noted that
  function name require variable lookups -- we distinguish those as they usually
  do not trigger computation.
\item \underline{$min(e)=\c{F(X)}$}: Function calls whose arguments may include
  variable references.

@@

\item \underline{$min(e)=\c{FUN}$}: Function definition accounts for
  \packageFunctionDefinitionSitesPercent of \eval sites. In addition the
  functions inside other expressions account for
  \packageGeneralizedFunctionDefinitionSitesPercent of sites, and give an
  estimate of the use of higher-order functions.
\item \underline{$min(e)=\c{\{\,\}}$}: Arbitrary-length statement blocks. These
  are larger expressions with, on average, \packageMinimizedoperationsjRnd
  instructions. The contents of the block are not inspected.
\item \underline{$min(e)=\c{model.frame}$}: Call to the \c{model.frame} function
  used to create statistical models. This subsume all other categories.

@@

%% Similarly, when a function is evaluated on some variable (\c{F(X)} minimized
%% expression), the percentage of non default, non parent environments is rather
%% high, \packageMinimizedpercentenvirnonparentd\%.

%% As a whole, variable access, either as the top call, or inside a function call,
%% possibly with accessing a slot, represents
%% \packageGeneralizedVarAccessSitePercent of the sites.

%% The minimized expression classification suggests that the presence of variables
%% in the expression is highly correlated to the specification of a specific
%% environment, and that \eval is used here for its power in precising the
%% evaluation environment.

Table~\ref{tab:minimizedexpressions} gives the most frequent forms; \#sites
(\%sites) is the number (ratio) of call sites receiving arguments of that form;
\#packages is the number of packages with that form; \#operations is the average
number of instructions performed by the interpreter when given that minimized
expression; and \%envir is the percentage of call sites for which the \c{envir}
argument is explicitly passed, and is not \c{parent.frame()}.

@@

A given \eval call site can be passed different forms at different invocations.
However, \packageNbOneMinimizedPercent of sites only see a single form.

\paragraph{Base}
The spread of forms in Base are similar to CRAN with a larger share of \c{V}, as
it is present in \baseMinimizedpropsitesa\% of the sites and \c{X} is a distant
second with \baseMinimizedpropsitesb\%.

\paragraph{Kaggle}
Kaggle is dominated by \c{V}, with \kaggleMinimizedpropsitesa\% of the
\kaggleNbCallSites sites.

\subsection{The environment of \eval}

The \c{envir} argument of \eval can be an actual environment, but also \c{Null},
a list, a dataframe, a pairlist or even an integer.
